### 2012-09-24 doc/prelim/document/chapters/linear_problem/linear_problem.tex
nonlinear updates

\section{Preliminaries}
\label{sec:linear_preliminaries}
We seek solutions of the general linear problem in the following form:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:,
  \label{eq:linear_problem}
\end{equation}
where $\ve{A} \in \mathbb{C}^{N \times N}$ is a matrix operator such
that $\ve{A} : \mathbb{C}^{N} \rightarrow \mathbb{C}^{N}$, $\ve{x} \in
\mathbb{C}^N$ is the solution vector, and $\ve{b} \in \mathbb{C}^N$ is
the forcing term. The solutions to Eq~(\ref{eq:linear_problem}) will
be generated by inverting $\ve{A}$ either directly or indirectly:
\begin{equation}
  \ve{x} = \ve{A}^{-1} \ve{b}
  \label{eq:linear_problem_solution}\:.
\end{equation}
In addition we can define the residual:
\begin{equation}
  \ve{r} = \ve{b} - \ve{A}\ve{x}\:,
  \label{eq:linear_residual}
\end{equation}
such that an exact solution $\ve{x}$ has been found when
$\ve{r}=\ve{0}$.  From the statement in
Eq~(\ref{eq:linear_problem_solution}) we can already place a
restriction on $\ve{A}$ by requiring that it be \textit{nonsingular},
meaning that we can in fact compute $\ve{A}^{-1}$. In this work we
will focus our efforts on approximately inverting the operator through
various means.

@@

General to both matrices and vectors, \textit{norms} are a mechanism
for collapsing objects of many elements to a single value. Per
LeVeque's text \citep{leveque_finite_2007}, the q-norm of a vector is defined
as:
\begin{equation}
  ||\ve{v}||_q = \Bigg[ \sum_{i=1}^N |v_i|^q \Bigg]^{1/q},\ \ve{v} \in
  \mathbb{C}^N\:,\ q \in \mathbb{Z}^+
  \label{eq:q_norm}
\end{equation}
where ${v_i}$ is the $i^{th}$ component of the vector. Depending on
the value chosen for $q$, local or global qualities of the vector may
be obtained. For example, $q=2$ provides the root of a quadrature sum
of all elements in the vector giving a global measure of the vector
while $q=\infty$ gives the maximum value in the vector, a local
quantity that does not give information regarding the other elements
in the vector.

We can also compute the norm of a matrix by inferring from the norm of
the vector on which it is operating. Per LeVeque, we search for a
constant that is equivalent to $||\ve{A}||$:
\begin{equation}
  ||\ve{A}\ve{x}|| \leq C ||\ve{x}||\:,
  \label{eq:matrix_norm_inequality}
\end{equation}
where the minimum value of $C$ that satisfies
Eq~(\ref{eq:matrix_norm_inequality}) is equivalent to $||\ve{A}||$
and is valid $\forall \ve{x} \in \mathbb{C}^N$. The general
definition in Eq~(\ref{eq:matrix_norm_inequality}) can be expanded in
simple terms for common norms including the infinity norm:
\begin{equation}
  ||\ve{A}||_{\infty} = \max_{1 \leq i \leq N} \sum^N_{j=1}|a_{ij}|\:,
  \label{eq:matrix_infinity_norm}
\end{equation}
and the 2-norm:
\begin{equation}
p  ||\ve{A}||_{2} = \sqrt{\rho(\ve{A}^T\ve{A})}\:,
  \label{eq:matrix_2_norm}
\end{equation}
where $\rho$ is the spectral radius as defined in
Eq~(\ref{eq:spectral_radius}).

@@

From Figure~\ref{fig:linear_projection_constraint} we then note that
the following geometric condition must hold:
\begin{equation}
  ||\tilde{\ve{r}}||_2 \leq ||\ve{r}_0||_2,\ \forall \ve{r}_0 \in
  \mathbb{C}^N\:,
  \label{eq:linear_min_res}
\end{equation}
meaning that the residual of the system will always be
\textit{minimized} with respect to the constraints.

Given this minimization condition for the residual, we can form the
outline of an iterative projection method. Consider a matrix $\ve{V}$
to form a basis of $\mathcal{K}$ and a matrix $\ve{W}$ to form a basis
of $\mathcal{L}$. As $\boldsymbol{\delta} \in \mathcal{K}$ by
definition in Eq~(\ref{eq:linear_projection_step}), then
$\boldsymbol{\delta}$ can instead be rewritten as:
\begin{equation}
  \boldsymbol{\delta} = \ve{V}\ve{y},\ \forall \ve{y} \in \mathbb{C}^N:\,
  \label{eq:linear_delta_projection}
\end{equation}
where $\ve{V}$ \textit{projects} $\ve{y}$ onto $\mathcal{K}$. From the
orthogonality constraint in Eq~(\ref{eq:linear_projection_constraint})
it then follows that:
\begin{equation}
  \ve{y} = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}_0\:,
  \label{eq:linear_constraint_projection}
\end{equation}
where here the projection onto $\mathcal{K}$ is constrained by the
projection onto $\mathcal{L}$. Knowing this, we can then outline the
following iteration scheme for a projection method:
\begin{subequations}
  \begin{gather}
    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
    \ve{y}^k = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}_0\:,\\
    \ve{x}^{k+1} = \ve{x}^k + \ve{V}\ve{y}^k\:,
  \end{gather}
  \label{eq:linear_projection_iteration}
\end{subequations}
where $\ve{V}$ and $\ve{W}$ are generated from the definitions of
$\mathcal{K}$ and $\mathcal{L}$ which are defined prior to each
iteration.

@@

Choosing GMRES as our model Krylov method, we are first tasked with
finding a projector onto the subspace. We seek an orthonormal basis
for $\mathcal{K}_m(\ve{A},\ve{r}_0)$ by an orthogonalization procedure
that is commonly based on, but not limited to, the \textit{Arnoldi}
recurrence relation. The Arnoldi procedure will generate an
orthonormal basis, $\ve{V}_m \in \mathbb{C}^{N \times m}$, via a
variant of the Gram-Schmidt procedure that re-applies the operator for
each consecutive vector, thus forming a basis that spans the subspace
in Eq~(\ref{eq:krylov_subspace}). Due to its equivalent
dimensionality, $m$, to that of the subspace, we will refer to such
recurrence relations as \textit{long recurrence relations}. Those
orthogonal projection procedures that have a dimensionality less than
$m$ will be refferred to as \textit{short recurrence relations}.  Once
$\ve{V}_m$ is found, per the constraint subspace definition it then
follows that its basis is defined as $\ve{W}_m = \ve{A}
\ve{V}_m$. Knowing the projections onto the search and constraint
subspaces, the GMRES iteration may be formulated as follows:
\begin{algorithm}[htpb!]
  \caption{GMRES Iteration}
  \label{alg:gmres}
  \begin{algorithmic}
    \State $\ve{r}_0 := \ve{b}-\ve{A}\ve{x}_0$
    \State $\beta := ||\ve{r}_0||_2$
    \State $\ve{v}_1 := \ve{r}_0 / \beta$
    \Comment{Create the orthonormal basis for the Krylov subspace}
    \For{$j = 1, 2, \cdots, m$}
    \State $h_{ij} \leftarrow \langle w_j,v_j \rangle$
    \State $w_j \leftarrow w_j - h_{ij}v_i$
    \EndFor
    \State $h_{j+1,j} \leftarrow ||w_j||_2$
    \State $v_{j+1} \leftarrow w_j / h_{j+1,j}$
    \Comment{Apply the orthogonality constraints}
    \State $\ve{y}_m \leftarrow argmin_y ||\beta \ve{e}_1 - \ve{H}_m\
    \ve{y}||_2 $
    \State $\ve{x}_m \leftarrow \ve{x}_0 + \ve{V_m} \ve{y}_m$
  \end{algorithmic}
\end{algorithm}

@@

\subsection{Parallel Vector Update}
\label{subsec:parallel_vec_update}
Parallel vector update operations arise from the construction of the
orthonormal basis and the application of the correction generated by
the constraints to the solution vector. Vector update operations are
embarassingly parallel in that they require no communication
operations to be successfully completed; all data operated on is
local. These operations are globally of the form:
\begin{equation}
  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in [1,N_g]
  \:,
  \label{eq:global_vector_update}
\end{equation}
and locally of the form:
\begin{equation}
  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in [1,N_l]
  \:,
  \label{eq:local_vector_update}
\end{equation}
where $\ve{y}$ and $\ve{x}$ are vectors of global size $N_g$, local
size $N_l$, and $a \in \mathbb{C}^N$. In order to avoid communcation,
the vectors $\ve{y}$ and $\ve{x}$ must have the same parallel
decomposition where each parallel process owns the same pieces of each
vector.

>>>>

\section{Preliminaries}
\label{sec:linear_preliminaries}
We seek solutions of the general linear problem in the following form:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:,
  \label{eq:linear_problem}
\end{equation}
where $\ve{A} \in \mathbb{R}^{N \times N}$ is a matrix operator such
that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
\mathbb{R}^N$ is the solution vector, and $\ve{b} \in \mathbb{R}^N$ is
the forcing term. The solutions to Eq~(\ref{eq:linear_problem}) will
be generated by inverting $\ve{A}$ either directly or indirectly:
\begin{equation}
  \ve{x} = \ve{A}^{-1} \ve{b}
  \label{eq:linear_problem_solution}\:.
\end{equation}
In addition we can define the residual:
\begin{equation}
  \ve{r} = \ve{b} - \ve{A}\ve{x}\:,
  \label{eq:linear_residual}
\end{equation}
such that an exact solution $\ve{x}$ has been found when
$\ve{r}=\ve{0}$.  From the statement in
Eq~(\ref{eq:linear_problem_solution}) we can already place a
restriction on $\ve{A}$ by requiring that it be \textit{nonsingular},
meaning that we can in fact compute $\ve{A}^{-1}$. In this work we
will focus our efforts on approximately inverting the operator through
various means.

@@

General to both matrices and vectors, \textit{norms} are a mechanism
for collapsing objects of many elements to a single value. Per
LeVeque's text \citep{leveque_finite_2007}, the q-norm of a vector is defined
as:
\begin{equation}
  ||\ve{v}||_q = \Bigg[ \sum_{i=1}^N |v_i|^q \Bigg]^{1/q},\ \ve{v} \in
  \mathbb{R}^N\:,\ q \in \mathbb{Z}^+
  \label{eq:q_norm}
\end{equation}
where ${v_i}$ is the $i^{th}$ component of the vector. Depending on
the value chosen for $q$, local or global qualities of the vector may
be obtained. For example, $q=2$ provides the root of a quadrature sum
of all elements in the vector giving a global measure of the vector
while $q=\infty$ gives the maximum value in the vector, a local
quantity that does not give information regarding the other elements
in the vector.

We can also compute the norm of a matrix by inferring from the norm of
the vector on which it is operating. Per LeVeque, we search for a
constant that is equivalent to $||\ve{A}||$:
\begin{equation}
  ||\ve{A}\ve{x}|| \leq C ||\ve{x}||\:,
  \label{eq:matrix_norm_inequality}
\end{equation}
where the minimum value of $C$ that satisfies
Eq~(\ref{eq:matrix_norm_inequality}) is equivalent to $||\ve{A}||$
and is valid $\forall \ve{x} \in \mathbb{R}^N$. The general
definition in Eq~(\ref{eq:matrix_norm_inequality}) can be expanded in
simple terms for common norms including the infinity norm:
\begin{equation}
  ||\ve{A}||_{\infty} = \max_{1 \leq i \leq N} \sum^N_{j=1}|a_{ij}|\:,
  \label{eq:matrix_infinity_norm}
\end{equation}
and the 2-norm:
\begin{equation}
p  ||\ve{A}||_{2} = \sqrt{\rho(\ve{A}^T\ve{A})}\:,
  \label{eq:matrix_2_norm}
\end{equation}
where $\rho$ is the spectral radius as defined in
Eq~(\ref{eq:spectral_radius}).

@@

From Figure~\ref{fig:linear_projection_constraint} we then note that
the following geometric condition must hold:
\begin{equation}
  ||\tilde{\ve{r}}||_2 \leq ||\ve{r}_0||_2,\ \forall \ve{r}_0 \in
  \mathbb{R}^N\:,
  \label{eq:linear_min_res}
\end{equation}
meaning that the residual of the system will always be
\textit{minimized} with respect to the constraints.

Given this minimization condition for the residual, we can form the
outline of an iterative projection method. Consider a matrix $\ve{V}$
to form a basis of $\mathcal{K}$ and a matrix $\ve{W}$ to form a basis
of $\mathcal{L}$. As $\boldsymbol{\delta} \in \mathcal{K}$ by
definition in Eq~(\ref{eq:linear_projection_step}), then
$\boldsymbol{\delta}$ can instead be rewritten as:
\begin{equation}
  \boldsymbol{\delta} = \ve{V}\ve{y},\ \forall \ve{y} \in \mathbb{R}^N:\,
  \label{eq:linear_delta_projection}
\end{equation}
where $\ve{V}$ \textit{projects} $\ve{y}$ onto $\mathcal{K}$. From the
orthogonality constraint in Eq~(\ref{eq:linear_projection_constraint})
it then follows that:
\begin{equation}
  \ve{y} = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}_0\:,
  \label{eq:linear_constraint_projection}
\end{equation}
where here the projection onto $\mathcal{K}$ is constrained by the
projection onto $\mathcal{L}$. Knowing this, we can then outline the
following iteration scheme for a projection method:
\begin{subequations}
  \begin{gather}
    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
    \ve{y}^k = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}_0\:,\\
    \ve{x}^{k+1} = \ve{x}^k + \ve{V}\ve{y}^k\:,
  \end{gather}
  \label{eq:linear_projection_iteration}
\end{subequations}
where $\ve{V}$ and $\ve{W}$ are generated from the definitions of
$\mathcal{K}$ and $\mathcal{L}$ which are defined prior to each
iteration.

@@

Choosing GMRES as our model Krylov method, we are first tasked with
finding a projector onto the subspace. We seek an orthonormal basis
for $\mathcal{K}_m(\ve{A},\ve{r}_0)$ by an orthogonalization procedure
that is commonly based on, but not limited to, the \textit{Arnoldi}
recurrence relation. The Arnoldi procedure will generate an
orthonormal basis, $\ve{V}_m \in \mathbb{R}^{N \times m}$, via a
variant of the Gram-Schmidt procedure that re-applies the operator for
each consecutive vector, thus forming a basis that spans the subspace
in Eq~(\ref{eq:krylov_subspace}). Due to its equivalent
dimensionality, $m$, to that of the subspace, we will refer to such
recurrence relations as \textit{long recurrence relations}. Those
orthogonal projection procedures that have a dimensionality less than
$m$ will be refferred to as \textit{short recurrence relations}.  Once
$\ve{V}_m$ is found, per the constraint subspace definition it then
follows that its basis is defined as $\ve{W}_m = \ve{A}
\ve{V}_m$. Knowing the projections onto the search and constraint
subspaces, the GMRES iteration may be formulated as follows:
\begin{algorithm}[htpb!]
  \caption{GMRES Iteration}
  \label{alg:gmres}
  \begin{algorithmic}
    \State $\ve{r}_0 := \ve{b}-\ve{A}\ve{x}_0$
    \State $\beta := ||\ve{r}_0||_2$
    \State $\ve{v}_1 := \ve{r}_0 / \beta$
    \Comment{Create the orthonormal basis for the Krylov subspace}
    \For{$j = 1, 2, \cdots, m$}
    \State $h_{ij} \leftarrow \langle w_j,v_j \rangle$
    \State $w_j \leftarrow w_j - h_{ij}v_i$
    \EndFor
    \State $h_{j+1,j} \leftarrow ||w_j||_2$
    \State $v_{j+1} \leftarrow w_j / h_{j+1,j}$
    \Comment{Apply the orthogonality constraints}
    \State $\ve{y}_m \leftarrow argmin_y ||\beta \ve{e}_1 - \ve{H}_m\
    \ve{y}||_2 $
    \State $\ve{x}_m \leftarrow \ve{x}_0 + \ve{V_m} \ve{y}_m$
  \end{algorithmic}
\end{algorithm}

@@

\subsection{Parallel Vector Update}
\label{subsec:parallel_vec_update}
Parallel vector update operations arise from the construction of the
orthonormal basis and the application of the correction generated by
the constraints to the solution vector. Vector update operations are
embarassingly parallel in that they require no communication
operations to be successfully completed; all data operated on is
local. These operations are globally of the form:
\begin{equation}
  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in [1,N_g]
  \:,
  \label{eq:global_vector_update}
\end{equation}
and locally of the form:
\begin{equation}
  \ve{y}[n] \leftarrow \ve{y}[n] + a * \ve{x}[n],\ \forall n \in [1,N_l]
  \:,
  \label{eq:local_vector_update}
\end{equation}
where $\ve{y}$ and $\ve{x}$ are vectors of global size $N_g$, local
size $N_l$, and $a \in \mathbb{R}^N$. In order to avoid communcation,
the vectors $\ve{y}$ and $\ve{x}$ must have the same parallel
decomposition where each parallel process owns the same pieces of each
vector.
